---
title: "On the origin of citations"
author: "Catherine Triandafillou"
date: "September 15, 2015"
output: 
  html_document:
    toc: true
    number_sections: true
    theme: cerulean
    highlight: espresso
    fig_width: 8
    fig_height: 8
    fig_caption: yes
    self_contained: false
---

# Load the data

```{r load_data}
counts_raw <- read.delim("data/counts-raw.txt")
counts_norm <- read.delim("data/counts-raw.txt")
```

# Data exploration

What's the distribution of authors in all articles of our dataset?

```{r author_histogram, echo=FALSE, fig.cap="Figure 1: Number of Authors per Article"}
hist(counts_raw$authorsCount, main = "Authors per paper", xlab="Number of authors")
```

```{r fb_histogram, echo=FALSE, fig.cap="Figure 2: Number of facebook shares per article"}
hist(counts_raw$facebookShareCount, main = "Facebook Shares per paper", xlab="Number of shares", xlim=c(0,30), breaks = 1000)
```

The average number of Facebook shares per paper in the dataset is `r mean(counts_raw$facebookShareCount)`

# dplyr

```{r, echo=FALSE}
library("dplyr")
```

```{r}
research <- filter(counts_raw, articleType == "Research Article")
```

```{r}
research_2006 <- filter(research, year == 2006,
                        facebookCommentCount > 0)
nrow(research_2006)
```

```{r}
research_2006_fb_tweet <- filter(research, year == 2006,
                        facebookCommentCount > 0 | backtweetsCount > 0)
nrow(research_2006)
```

```{r refer.to.columns}
#once research is declared in the dplyr function, all column names can be referred to natively
research_2006_fb_tweet_disease <- filter(research, year == 2006,
                        facebookCommentCount > 0 | backtweetsCount > 0,
                        grepl("Infectious Diseases", plosSubjectTags))
nrow(research_2006)
```

```{r dplyr.split.data}
#select is the dplyr command for selecting specific columns of our dataframe
article_info <- select(research, doi:authorsCount)
colnames(article_info)
```

```{r}
# contains option selects all columns with that pattern matched. Additional columns can be added by
# listing by hand, removed by listing with a minus sign.
metrics <- select(research, contains("Count"), -authorsCount, f1000Factor, wikipediaCites)
colnames(metrics)
```

```{r dplyr.slicing}
#list contents; select takes names or integers, slice takes only integers
head(select(research, 1:3))
slice(article_info, 1:3)
```

```{r dplyr.filter}
colnames(research)
low_cite <- filter(research, year < 2008,
                  pdfDownloadsCount > 1000,
                  mendeleyReadersCount > 15,
                  wosCountThru2011 < 10)
nrow(low_cite)
```

## Chaining commands with dplyr

pipe character %>%

```{r dplyr.pipes}
facebook_2006 <- research %>% filter(year == 2006) %>%
  select(contains("facebook"))
head(facebook_2006)

research %>% filter(year == 2006) %>%
  select(contains("facebook")) %>%
  nrow
```

arrange, works similar to function order; orders dataframe according to a numeric column

```{r dplyr.sorting}
research %>% arrange(desc(authorsCount), desc(wosCountThru2011)) %>%
  select(authorsCount, wosCountThru2011) %>%
  slice(1:10)
```

```{r}
research %>% arrange(desc(wosCountThru2011)) %>%
  select(title, wosCountThru2011) %>%
  slice(1:3)
```

```{r}
research %>% arrange(desc(authorsCount)) %>%
  select(authorsCount, title, journal, plosSubjectTags) %>%
  slice(1:3)
```

## Summarizing with dplyr

```{r dplyr.add.columns}
research <- mutate(research, weeksSincePublished = daysSincePublished / 7,
                   yearsSincePublished = weeksSincePublished / 52)
select(research, contains("Since")) %>% slice(1:10)
```

```{r}
summarize(research, plos_mean = mean(plosCommentCount),
          plos_sd = sd(plosCommentCount))
```

```{r dplyr.summarizing}
research %>% filter(journal == "pone", year == "2007") %>%
  summarize(plos_mean = mean(plosCommentCount),
          plos_sd = sd(plosCommentCount),
          num = n())
```

## Using groupby

```{r dplyr.groupby}
research %>% group_by(journal, year) %>%
  summarize(tweets_mean = mean(backtweetsCount))
```

```{r}
tweets_by_journal <- research %>% group_by(journal) %>%
  summarize(tweets_mean = mean(backtweetsCount),
            SEM = sd(backtweetsCount)/sqrt(n()))
tweets_by_journal
```

# ggplot2

## ggplot basics

```{r, echo=FALSE}
library("ggplot2")
```

```{r ggplot2.plot}
p <- ggplot(data = research, mapping = aes(x = pdfDownloadsCount,
                                           y = wosCountThru2011)) +
  geom_point(aes(color = journal, size = authorsCount, alpha = daysSincePublished))
p
```

```{r ggplot2.uniform.mapping}
#to assign something uniformly to a plot, don't use aes:
q <- ggplot(data = research, mapping = aes(x = pdfDownloadsCount,
                                           y = wosCountThru2011)) +
  geom_point(color = "red")
q
```

```{r}
# change the location of the aesthetic in order to change whether future layers are calculated on
# subsets or on the entire dataset.
p <- ggplot(data = research, mapping = aes(x = pdfDownloadsCount,
                                           y = wosCountThru2011,
                                           color = journal)) +
  geom_point() +
  geom_smooth()
p
```

```{r}
#challenge
p <- ggplot(data = research, mapping = aes(x = daysSincePublished,
                                           y = wosCountThru2011)) +
  geom_point(aes(color = journal), alpha = 0.5) +
  geom_smooth(color = "red")
p
```

```{r}
#problems with infinite values
p <- ggplot(data = research, mapping = aes(x = daysSincePublished,
                                           y = wosCountThru2011)) +
  geom_point(aes(color = journal)) +
  geom_smooth() +
  scale_x_log10() + scale_y_log10()
p
```

```{r ggplot2.change.axes}
#solutions
p <- ggplot(data = research, mapping = aes(x = log10(pdfDownloadsCount + 1),
                                           y = log10(wosCountThru2011 + 1))) +
  geom_point(aes(color = journal)) +
  geom_smooth() +
  #change the scaling so that the axes look logarithmic, change limits of the y axis
  scale_x_continuous(breaks = c(1, 3), labels = c(10, 1000)) +
  scale_y_continuous(breaks = c(1, 3), labels = c(10, 1000),
                     limits = c(1,3))
p
```


```{r ggplot2.colors.and.labels}
p + scale_color_grey()
p + scale_color_brewer(palette = "Dark2",
                       labels = 1:7, name = "PLOS")
```

```{r}
p <- ggplot(data = research, mapping = aes(x = sqrt(pdfDownloadsCount),
                                           y = sqrt(wosCountThru2011),
                                           color = journal,
                                           xlab = "Square Root of PDF Download Count",
                                           ylab = "Square Root of Citation Count")) +
  geom_point() +
  geom_smooth() +
  scale_color_brewer(palette = "Accent")
p
```

## subplots and facets

```{r ggplot2.facet}
p <- ggplot(research, aes(x = log10(pdfDownloadsCount + 1),
                          y = log10(wosCountThru2011 + 1))) +
  geom_point(aes(color = journal)) +
  geom_smooth() +
  scale_x_continuous(breaks = c(1, 3), labels = c(10, 1000)) +
  scale_y_continuous(breaks = c(1, 3), labels = c(10, 1000))

p + facet_wrap(~journal)
```

```{r ggplot.facetgrid}
research <- mutate(research, immuno = grepl("Immunology", plosSubjectTags))
# can also use %+% to update the dataframe after changing the base dataframe 'research'
p <- ggplot(research, aes(x = log10(pdfDownloadsCount + 1), y = log10(wosCountThru2011 + 1))) +
  geom_point(aes(color = journal)) +
  geom_smooth() +
  scale_x_continuous(breaks = c(1, 3), labels = c(10, 1000)) +
  scale_y_continuous(breaks = c(1, 3), labels = c(10, 1000))
p + facet_grid(journal~immuno)
```

## Different plot types

```{r}
p <- ggplot(research, aes(x = journal,
                          y = sqrt(wosCountThru2011))) +
  geom_violin()
p
```

```{r}
#recall tweets_per_journal again
tweets_bar <- ggplot(tweets_by_journal, aes(x = journal,
                                            y = tweets_mean)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin = tweets_mean - SEM, ymax = tweets_mean + SEM), width = 0.2)
tweets_bar
```

